---
name: "DS202 Lab 2 Team 8"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit the README.Rmd file -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Lab report #1

Follow the instructions posted at https://ds202-at-isu.github.io/labs.html for the lab assignment. The work is meant to be finished during the lab time, but you have time until Monday evening to polish things.

Include your answers in this document (Rmd file). Make sure that it knits properly (into the md file). Upload both the Rmd and the md file to your repository.  

All submissions to the github repo will be automatically uploaded for grading once the due date is passed.
Submit a link to your repository on Canvas (only one submission per team) to signal to the instructors that you are done with your submission.

TL;DR
```{r}
#install.packages('remotes')
#remotes::install_github("heike/classdata")
library(ggplot2)
``` 

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(classdata)

dim(ames)
str(ames)
```

## Types of Data
There are 16 columns with variables:

- Parcel ID: Useless, just an identifier
- Address: Kinda Useless
- Style: IMPORTANT, CATEGORICAL
- Occupancy: IMPORTANT, CATEGORICAL
- Sale Date: Maybe important for time series analysis. most likely QUANTITATIVE
- Sale Price: IMPORTANT, QUANTITATIVE (MAIN VARIABLE); RANGE: $0 - $22,000,000
- Multi Sale: CATEGORICAL (mostly NAs, watch out!) Useless in my opinion
- YearBuilt: Important for time series analysis. Most likely QUANTITATIVE; RANGE: 0 - 2022
- Acres: IMPORTANT, QUANTITATIVE; RANGE: 0 - 12.012
- TotalLivingArea (sf): IMPORTANT, QUANTITATIVE; RANGE: 0 - 6007
- Bedrooms: IMPORTANT, CATEGORICAL
- FinishedBsmtArea (sf): IMPORTANT, QUANTITATIVE; RANGE: 10 - 6496
- LotArea(sf): IMPORTANT, QUANTITATIVE; RANGE: 0 - 523228
- AC: IMPORTANT, CATEGORICAL
- FirePlace: IMPORTANT, CATEGORICAL
- Neighborhood: IMPORTANT, CATEGORICAL

**NOTE:**
- Most of the plots won't be done for max range of values because they might heavily skew the plots
- See [detailed visualization file called exploration_workspace.R in our repo](exploration_workspace.R) for all the various trends.\
We picked couple important ones for this document.\
DO NOT CLICK THE LINK FROM HTML AFTER KNITTING, THE LINK ONLY WORKS FOR README.md ON GITHUB

```{r, echo=FALSE}
ames['Bedrooms']
ggplot(ames, aes(Bedrooms))+
  geom_bar()
ggplot(ames, aes(x=`Bedrooms`, y=`Sale Price`)) + geom_point()
```

There are a lot of 2-5 bedroom houses

```{r, echo=FALSE}
ames %>% filter(`Sale Price` <= 1000000, `Sale Price` > 0) %>%
  ggplot(aes(x=`Sale Price`, fill=`FirePlace`)) + geom_histogram(binwidth=50000)
```

### Boxplots for `Sale Price`

Let's analyze boxplots of `Sale Price` for various types

#### By `Bedrooms`

```{r, message=FALSE, echo=FALSE}
# clean data
ames_clean_Style <- filter(ames, !is.na(Style))
ames_clean_Occupancy <- filter(ames, !is.na(Occupancy))
ames_clean_Bedrooms <- filter(ames, !is.na(Bedrooms) & Bedrooms != 0)
ames_clean_Neighborhood <- filter(ames, Neighborhood != "(0) None")

ggplot(data = filter(ames_clean_Bedrooms, `Sale Price` > 10000 & `Sale Price` <= 1000000), aes(`Sale Price`, y = factor(Bedrooms))) + geom_boxplot()

```

As you can see, the higher the bedrooms, the higher the sale price

#### By `Style`

```{r, message=FALSE, echo=FALSE}
ggplot(data = filter(ames_clean_Style, `Sale Price` > 10000 & `Sale Price` <= 1000000), aes(`Sale Price`, y = Style)) + geom_boxplot()
```

As you can see, 2 Story Brick has no outliers (every other type has outliers). It also has the highest IQR, possibly lots of variety in such houses for people to buy.

1 Story Frame has the highest range, however, most of the houses bought lie around $150,000 to $300,000

### Useful scatterplots and line plots

Let's now see some quantitative VS quantitative action

```{r, message=FALSE, echo=FALSE}
# clean data
ames_clean_scatterplot <- filter(ames, `Sale Price` > 10000 & `Sale Price` <= 1000000
                                     , !is.na(YearBuilt) & YearBuilt > 1850
                                     , !is.na(Acres) & Acres > 0 & Acres < 5
                                     , `TotalLivingArea (sf)` > 25 & `TotalLivingArea (sf)` < 3000
                                     , `FinishedBsmtArea (sf)` < 2000
                                     , !is.na(`LotArea(sf)`) & `LotArea(sf)` > 100 & `LotArea(sf)` < 15000
                                     , !is.na(Style) & !is.na(Occupancy) & !is.na(Bedrooms) & Bedrooms != 0 & Neighborhood != "(0) None")
```

#### Let's see `YearBuilt` VS `Sale Price`

```{r, message=FALSE, echo=FALSE}
ggplot(data = ames_clean_scatterplot, aes(x = YearBuilt, y = `Sale Price`)) + geom_point()
```

We see that overall trend seems to rise rapidly as we approach the 21st century.

We also see that there are some very cheap houses in the 21st century, maybe government offered good programs or there was a housing market crash.

Let's also see the trend more clearly by looking at average price per year

```{r, message=FALSE, echo=FALSE}
ames_clean_scatterplot %>%
  group_by(YearBuilt) %>%
  summarise(avg_sale_price = mean(`Sale Price`)) %>%
  ggplot(aes(x = YearBuilt, y = avg_sale_price)) + geom_line()
```

The rising trend near the 21th century is much more apparent now.

Let's also see how sale price varied for say, `Occupancy`

```{r, message=FALSE, echo=FALSE}
ames_clean_scatterplot %>%
  group_by(YearBuilt, Occupancy) %>%
  summarise(avg_sale_price = mean(`Sale Price`)) %>%
  ggplot(aes(x = YearBuilt, y = avg_sale_price, color = Occupancy)) + geom_line()
```

Townhouse trend started after 1960s. All types exhibit rising trend, however, two family duplex was expensive around 1920s but fell off later, rose a little bit but relatively stable.

#### Let's see `LotArea(sf)` VS `Sale Price`

Let's also throw in a category `Occupancy` while we're at it

```{r, message=FALSE, echo=FALSE}
ames_clean_scatterplot %>%
  group_by(Occupancy, `LotArea(sf)`) %>%
  summarise(avg_sale_price = mean(`Sale Price`)) %>%
  ggplot(aes(x = `LotArea(sf)`, y = avg_sale_price, color = Occupancy)) + geom_point()
```

We notice interesting patterns.

Townhouses usually have a smaller area but price rises rapidly as area increases.

Two Family conversion has a slight increase with area.

Single family/owner occupied is clearly the most popular and is usually larger area than townhouses and has wide price range.


#Varun:
Question 1)

# Display the structure of the Ames dataset
```{r}
str(ames)
```
Question 2)

I am interested in sale price becuase I believe it is a key variable for real estate analysis.
```{r}
# Summary of Sale Price
summary(ames$`Sale Price`)

```

I believe the numbers show that most houses in Ames sell for around 170,900, which seems like the normal price for a house there. But there are some houses that sold for a lot more, making the average price look really high at 1,017,479. Also, some houses are listed with a sale price of 0, which is strange and might be a mistake or something special about those sales. So, I think if we don't count those unusual 0 and super high prices, we'd get a clearer picture of what most houses really cost in Ames.


Question 3) 
```{r}
# Histogram for Sale Price
ggplot(ames, aes(x=`Sale Price`)) +
  geom_histogram(binwidth = 20000, fill = "steelblue", color = "black") +
  labs(title = "Histogram of Sale Prices", x = "Sale Price", y = "Frequency") +
  theme_minimal()

```
```{r}
ggplot(ames, aes(x=`Sale Price`)) +
  geom_histogram(binwidth = 50000, fill = "dodgerblue", color = "black") +
  xlim(c(0, 1000000)) +  # Limiting x-axis to focus on the common range
  labs(title = "Distribution of Sale Prices", x = "Sale Price", y = "Count") +
  theme_minimal()

```
```{r}
ggplot(ames, aes(x=`Sale Price`)) +
  geom_density(fill="salmon", alpha=0.5) +
  xlim(c(0, 1000000)) +  # Focus on the common price range
  labs(title = "Density of Sale Prices", x = "Sale Price", y = "Density") +
  theme_minimal()

```
```{r}
ggplot(ames, aes(x=`Sale Price`)) +
  stat_ecdf(geom = "step", color="steelblue") +
  xlim(c(0, 1000000)) +  # Focus on the common price range
  labs(title = "Cumulative Distribution of Sale Prices", x = "Sale Price", y = "Cumulative Frequency") +
  theme_minimal()


```

Main Variable: Sale Price
Range of Sale Price
The range of "Sale Price" was already identified in the summary provided earlier, with a minimum of 0 and a maximum of 20,500,000. However, the presence of 0 values suggests some transactions might not represent actual sales.



```{r}
# Bar chart for Bedrooms
ggplot(ames, aes(x=factor(Bedrooms))) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Distribution of Bedrooms", x = "Number of Bedrooms", y = "Count") +
  theme_minimal()
```


4)

Range of Total Living Area
```{r}
summary(ames$`TotalLivingArea (sf)`)

```
The statistics for the 'TotalLivingArea (sf)' variable in the dataset reveal a broad range, with a minimum of 0 square feet, suggesting possible data errors, and a maximum of 6,007 square feet, indicating a diverse property size spectrum. The first quartile is at 1,095 square feet, the median is 1,460 square feet, and the third quartile is at 1,792 square feet, with an average living area of approximately 1,507 square feet. This average, being higher than the median, hints at a right-skewed distribution with some exceptionally large properties. Additionally, there are 447 entries without recorded living area data, which could impact the analysis and might necessitate data cleaning or imputation.

```{r}
range(classdata::ames$Bedrooms, na.rm = TRUE)

bar_chart_bedrooms <- ggplot(classdata::ames, aes(x=as.factor(Bedrooms))) +
  geom_bar(fill="steelblue", color="black", alpha=0.7) +
  labs(title="Distribution of Number of Bedrooms", x="Number of Bedrooms", y="Frequency") +
  theme_minimal()

print(bar_chart_bedrooms)
```

I think there are oddities in the barchart between frequency and number of bedrooms.I believe there are a couple of notable oddities. Firstly, the bar for 0 bedrooms is peculiar, as typically a property would have at least one bedroom; this suggests either a data entry error or a unique type of dwelling that is not meant for typical residence, like a studio. Secondly, there's a bar for 10 bedrooms which is quite unusual for standard homes, indicating that this could be a large multi-family dwelling or a data anomaly. Lastly, the 'NA' category is present, implying that there are properties with unspecified bedroom counts, which could impact the analysis if these entries are significant in number. It might be necessary to investigate these anomalies to ensure accurate data analysis.


```{r}
boxplot_sale_price_vs_bedrooms <- ggplot(classdata::ames, aes(x=as.factor(Bedrooms), y=`Sale Price`)) +
  geom_boxplot(fill="steelblue", color="black", alpha=0.7) +
  labs(title="Boxplot of Sale Prices by Number of Bedrooms", x="Number of Bedrooms", y="Sale Price") +
  theme_minimal()

print(boxplot_sale_price_vs_bedrooms)
```

The boxplot depicting sale prices by the number of bedrooms shows a wide variation in one-bedroom properties and outliers in several categories, suggesting diverse property types or data inconsistencies. The 'NA' category indicates missing bedroom information, and the presence of a ten-bedroom category could reflect exceptionally large residences. The lack of a clear price trend with increasing bedrooms and the scientific notation on the y-axis suggests a vast range of sale prices.


After reviewing all the graphs, it looks like the sale prices of houses vary a lot, especially for homes with just one bedroom. Also, there are some homes with a lot of bedrooms that have prices that are really different from the rest. It's not always true that more bedrooms mean a more expensive house. There's also some information missing for some houses (like how many bedrooms they have), which makes it harder to see clear patterns.

#Fun

```{r}
ggplot(ames, aes(x=YearBuilt, y=`Sale Price`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Sale Price vs. Year Built", x = "Year Built", y = "Sale Price") +
  theme_minimal()

ggplot(ames, aes(x=Neighborhood, y=`Sale Price`)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Sale Price by Neighborhood", x = "Neighborhood", y = "Sale Price")

```
This graph shows how much houses are sold for in different neighborhoods. Most places have homes sold for around the same prices, but there's one neighborhood on the right where houses are much more expensive. This neighborhood also has houses with a big range of prices, from kind of expensive to super expensive. This probably means this is a fancy area.

# MACHINE LEARNING!!!

```{r, message=FALSE, echo=FALSE}
# Get the data ready 
ames_clean <- subset(ames, select = -c(`Parcel ID`, Address, `Sale Date`, `Multi Sale`, `FinishedBsmtArea (sf)`))
ames_clean <- filter(ames_clean, `Sale Price` > 100000 & `Sale Price` <= 1000000
                             , !is.na(YearBuilt) & YearBuilt > 1850
                             , !is.na(Acres) & Acres > 0 & Acres < 5
                             , `TotalLivingArea (sf)` > 25 & `TotalLivingArea (sf)` < 3000
                             #, `FinishedBsmtArea (sf)` < 2000 # Don't uncomment cuz we deleted this column
                             , !is.na(`LotArea(sf)`) & `LotArea(sf)` > 100 & `LotArea(sf)` < 15000
                             , !is.na(Style) & !is.na(Occupancy) & !is.na(Bedrooms) & Bedrooms != 0 & Neighborhood != "(0) None")
ames_clean$AC <- factor(ames_clean$AC)
ames_clean$FirePlace <- factor(ames_clean$FirePlace)
ames_clean$Bedrooms <- factor(ames_clean$Bedrooms)
ames_clean <- ames_clean %>% mutate(
  PriceClass = NA,
  PriceClass = if_else(`Sale Price` > 25000  & `Sale Price` <= 125000,
                   1, PriceClass),
  PriceClass = if_else(`Sale Price` > 125000 & `Sale Price` <= 375000,
                   2, PriceClass),
  PriceClass = if_else(`Sale Price` > 375000 & `Sale Price` <= 500000,
                   3, PriceClass),
  PriceClass = if_else(`Sale Price` > 500000 & `Sale Price` <= 600000,
                   4, PriceClass),
  PriceClass = if_else(`Sale Price` > 600000 & `Sale Price` <= 800000,
                   5, PriceClass),
  PriceClass = if_else(`Sale Price` > 800000 & `Sale Price` <= 1000000,
                   6, PriceClass)
)
ames_clean$PriceClass <- factor(ames_clean$PriceClass)
ames_test <- ames_clean
```

### Data exploration/visualization for ML
  
Let's explore data for ML purposes
    
```{r, message=FALSE, echo=FALSE}
ames_test %>% keep(is.numeric) %>% gather() %>%
ggplot(aes(x=value)) +
geom_histogram(fill="steelblue", alpha=.7) +
theme_minimal() +
facet_wrap(~key, scales="free")

cor(ames_test %>% keep(is.double)) %>%
as.data.frame %>% mutate(var2=rownames(.)) %>%
pivot_longer(!var2, values_to = "value") %>%
ggplot(aes(x=name,y=var2,fill=abs(value),label=round(value,2))) +
geom_tile() + geom_label() + xlab("") + ylab("") +
ggtitle("Correlation matrix of our predictors") +
labs(fill="Correlation\n(absolute):")

ames_test %>% select_if(is.double) %>%
pivot_longer(!`Sale Price`, values_to = "value") %>%
ggplot(aes(x=value, y=`Sale Price`)) +
geom_point() +
theme_minimal() +
facet_wrap(~name, scales="free")

function_CAT_SalePrice <- function(x){
  x_CAT_n_1QUANT <- data.frame(matrix(ncol = 0, nrow = dim(x)[1]))
  x_CAT_n_1QUANT <- cbind(x_CAT_n_1QUANT, x['Sale Price'], select_if(x, is.factor))
  return (x_CAT_n_1QUANT)
}
ames_test %>% function_CAT_SalePrice() %>%
  pivot_longer(!`Sale Price`, values_to = "value") %>%
  ggplot(aes(x=`Sale Price`, fill=value)) + # can use x=log(`Sale Price`) to make distribution more normal
  geom_histogram(show.legend = FALSE) +
  facet_wrap(~name, scales="free")

ames_test %>% function_CAT_SalePrice() %>%
  pivot_longer(!`Sale Price`, values_to = "value") %>%
  ggplot(aes(y=`Sale Price`, x = value)) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(~name, scales="free")
```

We notice not much correlations. Also varying behavior based on various factors. So simple linear regression or clustering won't work.

### Model training

Let's use Random Forest Regression as decision trees certainly help determine the important features.

```{r, message=FALSE, echo=FALSE}
# machine learning library
library(caret)

split <- sample(1:nrow(ames_test), as.integer(0.7*nrow(ames_test)), F)

train <- ames_test[split, ]
test  <- ames_test[-split,]

x_train <- train[,names(train) != "Sale Price"]
x_test  <- test [,names(test ) != "Sale Price"]

y_train <- unlist(train[,"Sale Price"])
y_test <- unlist(test[,"Sale Price"])

mod <- caret::train(x_train, y_train, method="rf",
                tuneGrid = expand.grid(mtry = seq(1,ncol(x_train),by=1)),
                trControl = trainControl(method="cv", number=2, verboseIter = T))
```

After model is trained, let's see most important features, RMSE and our prediction VS actual plots.

```{r, message=FALSE, echo=FALSE}
plot(varImp(mod), main="Feature importance of random forest model on training data")

print(mod)
predictions <- predict(mod, newdata = x_test)
dff <- cbind(predictions, y_test)
# predicted vs real plot
as.data.frame(dff) %>% ggplot(aes(x = predictions, y = y_test)) + geom_point() + geom_abline(aes(intercept = 0 , slope = 1))
```


Most important feature is price class followed by neighborhood which makes sense.

We see our model predicts quite accurately. Will optimize further for finer tuning later. 
